Merging merges, more or less

### 2015-03-04 2015mergingmerges/section/02-Machines.tex
merge section

>>>>

Note that @Out@, @OutDone@, @If@ and @Update@ are also annotated with the name of the output channel.
While a single combinator will only have one output, after merging multiple machines together there could be multiple outputs.
Each combinator has its own local mutable state, accessible by the functions given to @Update@, @Out@ and @If@.
These machine state types must be annotated with the name of the output channel to designate which local mutable state, if any.

### 2015-03-04 2015mergingmerges/section/03-Access.tex
merge section

%!TEX root = ../Main.tex
\section{Access patterns}
\label{s:Access}

The key insight is to treat combinators' access patterns as regular expressions over the names of streams.
When an input stream is mentioned in a regular expression, it signifies that a blocking read must be made from that stream before the combinator can proceed.
When output, written @E@ for emit, is mentioned, a blocking write is made to the output.
We associate these regular expression access patterns with a combinator, process or network, such that for any given set of inputs, there exists an execution trace that satisfies the regular expression:
\[ \forall g : @Graph@.~ \forall is : @Inputs@.~ \exists t : @Trace@~g~is.~
    \flatten(t) \in \regex(g) \]

Where $\flatten : @Trace@~g~i \to [@Index@~n]$ converts an execution trace of a graph to the list of indices as they fired, and $\regex : @Graph@ \to @Regex@~(@Index@~n)$ retrieves the access pattern of a given graph.

For example, @map a@ reads from @a@ at every step, then emits a value @E@, and repeats until no more data remains in @a@.
We define the operator @:~@ as ``has access pattern''.
\begin{code}
map     a   :~ (aE)*
filter  a   :~ (aE?)*
zip     a b :~ (abE)*
append  a b :~ (aE)*(bE)*
merjoin a b :~ ((aE)*(bE)*)*
\end{code}
The access pattern of the merge join (@merjoin@) also illustrates its relationship with append and segmented append.
These access patterns be composed together to apply to whole graphs:

\begin{code}
append (map a) (filter b) :~ (aE)*(bE?)*
append (zip a b) (zip a b) :~ (abE)*(abE)*
\end{code}
\TODO{but I don't have a good definition for this yet}

\subsection{Traces}
\TODO{Define what a trace is}.

We define filter functions over traces and regular expressions, which, given two graphs and the trace or acess pattern of one, filters it to the intersection of the names.
\[ \filterT(g', t) = \filter (\in flatten(g')) t \]
\[ \filterR(g', r) = \filter (\in flatten(g')) r \]
It also holds that
\[ \forall rx. \forall s. \forall w. w \in rx \implies \filter(\in s) w \in \filter(\in s) rx\]

We say that two traces are \emph{affable} if they agree on common parts:
\[ \affable(t : @Trace@~g~is, t' : @Trace@~g'~is) = \filterT(g', t) == \filterT(g, t') \]

For example, any traces of @map a@ and @zip a b@ would be affable, since after filtering out the @b@s both traces would be equal.

Two graphs are fusible if, for all possible inputs, there exist affable traces:
\begin{align*}
\mathit{fusible}(g, g')
        &=  \forall is : @Inputs@~(g \cap g').  \\
        &   \exists t  : @Trace@~g~is.          \\
        &   \exists t' : @Trace@~g'~is.         \\
        &   \affable(t, t')                     %
\end{align*}

Indeed, if we have two graphs, $\forall g~g' : @Graph@.$
\begin{eqnarray*}
 & \filterR(\regex(g)) = \filterR(\regex(g')) \\
 \implies & \forall w. w \in \filterR(\regex(g)) \iff w \in \filterR(\regex(g')) \\
 \implies & \forall is. \exists t : @Trace@~g~is. \exists t' : @Trace@~g'~is. \\
          & \filterT(\flatten(t)) \in \filterR(\regex(g)) \\
          & \wedge~\filterT(\flatten(t')) \in \filterR(\regex(g')) \\
          & \wedge~\filterT t = \filterT t' \\
 \implies & \affable(t, t') \\
 \implies & \mathit{fusible}(g, g') \\
\end{eqnarray*}

>>>>

%!TEX root = ../Main.tex
\section{Merging machines}
\label{s:Merging}

This section describes how machines can be merged together to form a single machine that computes both.

The idea is to create a new machine with the product of the two states; at each state, one or the other of the machines may be able to take a step.
If so, we update that machine's state and leave the other one as-is.
For shared inputs or outputs, both machines must take the steps at the same time: if they both pull from the same input, they must pull at the same time.
Similarly, if one machine produces an input and the other reads it, the first machine can only produce its output if the second is pulling on it.
If neither machine can make a move, the program is disallowed.
For example, if the first machine is trying to read from a shared input while the second tries to read from the first machine's input, executing it may require buffering the input until the second machine catches up.

However, the above explanation is too strict and outlaws programs such as @zip xs ys@ merged with @zip ys xs@, as neither machine can progress until the other does, leading to deadlock.
To resolve this, we add to the resulting machine's state a set of pending events for each machine.
The first machine may read from a shared input as long as there is not already a ``read'' event pending for the other machine to deal with.
Reading from a shared input adds a pending ``read'' event to the other machine, and the other machine may then skip past a @Pull@ for that source, as it is known that the other machine has already pulled on it.

\subsection{General merging}
Talk about merging with
\begin{code}
data Event
 = Read n
 | Finished n

merge :: Machine l1 -> Machine l2
      -> Machine (l1, l2, Set Event)
\end{code}

\subsection{Producer-consumer}
The general case above is sufficient, but in some cases can generate larger machines than necessary.
Talk about special case for producer-consumers.

\begin{code}
data Which = LeftExec | RightExec

mergeV :: Machine l1 -> Machine l2
       -> Machine (l1, l2, Which)
\end{code}
